Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French

### 2019-06-03 article/article.tex
Working on article, huge enhancements of some metrics

\title{Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French and Latin}
\author[1]{Thibault Clérice}
\affil[1]{École nationale des Chartes, France} 
\affil[2]{Université Lyon 3, France}

@@

\begin{document}

@@

\keywords{convolutional network; scripta continua; tokenization; Latin; Old French; word segmentation}

@@

% To Read : Stutzmann article.

Tokenization of space-less strings is a task that is specifically difficult for computer when compared to "whathumancando". \textit{Scripta continua} is a writing phenomenon where words would not be separated by spaces and it appears to have disappeared around the 8th century (see \citet{zanna1998lecture}). Although, spacing can be somewhat erratic in later centuries (\textit{cf.} Figure \ref{fig:4lines}). In the context of text mining of HTR or OCR output, lemmatization and tokenization of medieval western languages can be a pre-processing step for further research to sustain analyses such as authorship attribution \textbf{CITE JBCAMPS ?}.

@@

@@

The model is based on traditional text input encoding where each character is transcoded to an index. Output of the model is a mask that needs to be applied to the input: in the mask, characters are classified either as word boundary or word content.

\begin{table}[]
\centering
\begin{tabular}{@{}ll@{}}
\hline
                       & \textbf{Sample}           \\  \hline
\textbf{Input  String} & \texttt{Ladamehaitees'enparti}     \\
\textbf{Mask   String} & \texttt{xSxxxSxxxxxSxxxSxxxxS}     \\
\textbf{Output String} & \texttt{La dame haitee s'en parti} \\ \hline
\end{tabular}
  \caption{Input, mask and human-readable output generated by the model. x are WC and S are WB}
  \label{lst:input_output_example}
\end{table}

For evaluation purposes, and to reduce the number of input classes, we propose both a lower-case normalization and a "reduction to the ASCII character set" feature (fr. \ref{fig:normalization}). On this point, a lot of issues were found with transliteration of MUFI characters that were part of the original datasets, as they are badly interpreted by the \texttt{unidecode} python package. Indeed, \texttt{unidecode} will simply remove characters it does not understand. An addendum to said package that would transcribe ligatures, abbreviations, etc. would probably be a good feature.

% Need a better system that this/
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{inputs.png}
  \caption{ Different possibilities of pre-processing. In the last, the first character (LATIN ABBREVIATION SIGN SMALL ET WITH STROKE) is lost by \texttt{unidecode}}
  \label{fig:normalization}
\end{figure}

@@

Each model we have used is composed of an Encoder, generally built around an embedding layer, and the same Linear Classifier. The encoders tested are:

\begin{itemize}
  \item LSTM encoder with hidden cell
  \item Convolutional (CNN) encoder with position embeddings 
  \item Convolutional (CNN) encoder without position embeddings % Currently not really because it's part of the network, but it should be for the release of the article
\end{itemize}

@@

Datasets are transcription from manuscripts with unresolved abbreviation coming from different projects

\begin{itemize}
    \item \textbf{Old French} based on \citet{8269990}, \citet{pinche:hal-01628533}, \citet{jean_baptiste_camps_2019_2630574}, \citet{bfmmss}, and transcription from the TNAH master at the École Nationale des Chartes.
    \begin{itemize}
        \item 193,734 training examples;
        \item 23,581 validation examples;
        \item 25,512 test examples
    \end{itemize}
\end{itemize}

@@

\begin{figure}[h!]
  \begin{center}
    \begin{tikzpicture}
      \begin{axis}[
          width=\linewidth, % Scale the plot to \linewidth
          grid=major, 
          grid style={dashed,gray!30},
          xlabel=Epoch, % Set the labels
          ylabel=Accuracy,
          legend style={at={(0.5,-0.2)},anchor=north},
          x tick label style={rotate=90,anchor=east}
        ]
        \addplot table[x=Epoch,y=CNN Train,col sep=comma] {accuracies.csv}; 
        \addplot table[x=Epoch,y=CNN No Pos,col sep=comma] {accuracies.csv}; 
        \addplot table[x=Epoch,y=CNN lower,col sep=comma] {accuracies.csv}; 
        \addplot table[x=Epoch,y=CNN norm,col sep=comma] {accuracies.csv}; 
        \legend{CNN Train,CNN without Position embedding, CNN with lowercase transform, CNN with ASCIIification}
      \end{axis}
    \end{tikzpicture}
    \caption{Training Accuracy over 50 epochs}
  \label{fig:accuracy}
  \end{center}
\end{figure}

% Include evaluation on test sets

@@

The following input was not seen during training, dev or test, and is part of a different corpus :

@@

\section{Annex 1}
Pellentesque dignissim ultrices fringilla. Vivamus eu luctus ante, vel bibendum magna. Curabitur elit purus, tincidunt non dui
vitae, elementum bibendum neque. Curabitur ullamcorper sit amet justo at hendrerit. Fusce ut arcu imperdiet nibh mollis
tempus a aliquet tellus. Quisque pharetra cursus nisi, vel lobortis ante consectetur et. Vivamus sed congue neque. Proin
pellentesque risus nec dui consequat rutrum. Vestibulum nunc diam, placerat quis auctor vel, faucibus non justo. Etiam
dictum purus neque. Phasellus imperdiet mauris ligula, eu laoreet nisi elementum ut. Sed sed porta massa. Aenean faucibus
risus ultrices ornare porta. Quisque faucibus ante a tincidunt vestibulum. Lorem ipsum dolor sit amet, consectetur adipiscing
elit.

>>>>

\title{Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French}
\author[1]{Thibault Clérice}
\affil[1]{École nationale des Chartes, France} 
\affil[2]{Université Lyon 3, France}

@@

\pgfplotsset{compat=1.15}
\begin{document}

@@

\keywords{convolutional network; scripta continua; tokenization; Old French; word segmentation}

@@

% To Read : Stutzmann article.
% Parler plus de la variation orthographique de langue

Tokenization of space-less strings is a task that is specifically difficult for computer when compared to "whathumancando". \textit{Scripta continua} is a writing phenomenon where words would not be separated by spaces and it appears to have disappeared around the 8th century (see \citet{zanna1998lecture}). Never the less, spacing can be somewhat erratic in later centuries writings, as show by Figure \ref{fig:4lines}, a document from the 13th century. In the context of text mining of HTR or OCR output, lemmatization and tokenization of medieval western languages can be a pre-processing step for further research to sustain analyses such as authorship attribution \textbf{CITE JBCAMPS ?}.

@@

Indeed, while Chinese's issue seems to lie in the decomposition of relatively fix characters, Old French or medieval latin present heavy variation of spelling. In \citet{camps_pandora}, Camps notes, in the same corpus, the existence of not less than 29 spelling of the word \textit{cheval} (horse in Old and Modern French) whose apparition counts span from 3907 to 1\footnote{These are \textit{cheval}, \textit{chevaus}, \textit{cheual}, \textit{ceval}, \textit{chevals}, \textit{cevaus}, \textit{chival}, \textit{ceual}, \textit{cheuaus}, \textit{cevals}, \textit{chaval}, \textit{chivaus}, \textit{chiual}, \textit{chevas}, \textit{cheuals}, \textit{chiuaus}, \textit{ceuaus}, \textit{chevaul}, \textit{chiuau}, \textit{chivals}, \textit{chevau}, \textit{kevaus}, \textit{chavaus}, \textit{cheuas}, \textit{keval}, \textit{cheua}, \textit{cheuau}, \textit{cheva}, \textit{chiuals}}. This  makes a dictionary approach rather difficult as it would rely on a high number of different spelling and makes the computation highly complex.

@@

The model is based on traditional text input encoding where each character is transcoded to an index. Output of the model is a mask that needs to be applied to the input: in the mask, characters are classified either as word boundary or word content (\textit{cf.} Table \ref{lst:input_output_example}.

\begin{table}
\centering
\begin{tabular}{@{}ll@{}}
\hline
                       & \textbf{Sample}           \\  \hline
\textbf{Input  String} & \texttt{Ladamehaitees'enparti}     \\
\textbf{Mask   String} & \texttt{xSxxxSxxxxxSxxxSxxxxS}     \\
\textbf{Output String} & \texttt{La dame haitee s'en parti} \\ \hline
\end{tabular}
  \caption{Input, mask and human-readable output generated by the model. x are WC and S are WB}
  \label{lst:input_output_example}
\end{table}

For evaluation purposes, and to reduce the number of input classes, we propose two options for data transcoding: a lower-case normalization and a "reduction to the ASCII character set" feature (fr. \ref{fig:normalization}). On this point, a lot of issues were found with transliteration of medieval paelographic characters that were part of the original datasets, as they are badly interpreted by the \texttt{unidecode} python package. Indeed, \texttt{unidecode} will simply remove characters it does not understand. I built a secondary package named \texttt{mufidecode} (\citet{thibault_clerice_2019_3237731}) which precedes unidecode equivalency tables when the data is known of the Medieval Unicode Font Initiative (MUFI, \citet{mufi}).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{carbon.png}
  \caption{Different possibilities of pre-processing. The option with join=False was kept, as it keeps abbreviation marked as single characters. Note how \texttt{unidecode} loses the P WITH BAR}
  \label{fig:normalization}
\end{figure}

@@

Aside from normalizations of the input and output, three different structure of models were tested. Every model is composed by one encoder described below and one Linear Classifier which classifies into 5 classes : Start of Sentence (= SOS), End of Sentence (= EOS), Padding (= PAD), Masked Token (= Word Content), Space (= Word Boundary). For final scores, SOS, EOS and PAD were ignored.

The encoders are the following (configurations in parenthesis):

\begin{itemize}
  \item LSTM encoder with hidden cell (Embedding (512), Dropout(0.5), Hidden Dimension (512), Layers(10))
  \item Convolutional (CNN) encoder with position embeddings (Embedding (256), Embedding(Maximum Sentence Size=150), Kernel Size (5), Dropout(0.25), Layers (10))
  \item Convolutional (CNN) encoder without position embeddings (Embedding (256), Kernel Size (5), Dropout(0.25), Layers (10))
\end{itemize}

@@

Datasets are transcription from manuscripts with unresolved abbreviation coming from different projects. The \textbf{Old French} is based on \citet{8269990}, \citet{pinche:hal-01628533}, \citet{jean_baptiste_camps_2019_2630574}, \citet{bfmmss}, and \citet{tnah_transcription}. It contains

\begin{itemize}
    \item 193,734 training examples;
    \item 23,581 validation examples;
    \item 25,512 test examples
\end{itemize}

The input was generated by grouping at least 2 words and a maximum of 8 words together per sample. On a probability of 0.2, noise character could be added (noise character was set to DOT ('.')) and some words were kept randomly from a sample to another on a probability of 0.3 and a maximum number of word kept of 1. If a minimum size of 7 characters was not met in the input sample, another word would be added to the chain. A maximum input size of 100 was kept. The results corpora should be varied in sizes as shown by \ref{fig:word_sizes}. The corpora is composed by 193 different characters when not normalized, in which some MUFI characters appears few hundred times \ref{tab:mufi_examples}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{length.png}
  \caption{Distribution of word size over the train, dev and test corpora}
  \label{fig:word_sizes}
\end{figure}

\begin{table}
\begin{tabular}{llll}
\hline
                                                   & Train dataset & Dev dataset & Test dataset \\ \hline
TIRONIAN SIGN ET                                   & 4367          & 541         & 539          \\
CON                             & 508           & 70          & 76           \\
P WITH STROKE THROUGH DESCENDER & 580           & 69          & 84           \\ \hline
\end{tabular}
  \caption{Examples of some MUFI characters distributions}
  \label{tab:mufi_examples}
\end{table}

@@

The training parameters was 0.00005 in learning rate for each CNN model and 0.001 for the LSTM one, and 64 in batch sizes. Training reached a plateau fairly quickly for each model (\textit{cf.} \ref{fig:loss}). Each model except LSTM reached a really low loss and a high accuracy on the test set (\textit{cf.} \ref{tab:scores})

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \begin{axis}[
          width=\linewidth, % Scale the plot to \linewidth
          grid=major,
          grid style={dashed,gray!30},
          xlabel=Epoch, % Set the labels
          ylabel=Accuracy,
          legend style={at={(0.5,-0.2)},anchor=north},
          x tick label style={rotate=90,anchor=east}
        ]
        \addplot table[x=Epoch,y=CNN 1,col sep=comma] {accuracies.csv};
        \addplot table[x=Epoch,y=CNN L,col sep=comma] {accuracies.csv};
        \addplot table[x=Epoch,y=CNN w/o P,col sep=comma] {accuracies.csv};
        \addplot table[x=Epoch,y=CNN N,col sep=comma] {accuracies.csv};
        \addplot table[x=Epoch,y=CNN L N,col sep=comma] {accuracies.csv};
        \legend{CNN, CNN L, CNN P, CNN N, CNN L N}
      \end{axis}
    \end{tikzpicture}
    \caption{Training Loss (Cross-entropy) until plateau was reached. N = normalized, L = Lower, P = no position embedding. LSTM was removed as it did not go below 0.65}
  \label{fig:loss}
  \end{center}
\end{figure}

\begin{table}
\centering
\begin{tabular}{lllll}
\hline
Model   & Accuracy & Precision & Recall & FScore \\ \hline
CNN     & 0.991    & 0.985     & 0.990  & 0.987  \\
CNN L   & 0.991    & 0.979     & 0.990  & 0.985  \\
CNN P   & \textbf{0.993}    & \textbf{0.990}& \textbf{0.991}  & \textbf{0.990}  \\
CNN N   & 0.991    & 0.987     & 0.988  & 0.988  \\
CNN L N & 0.992    & 0.988     & 0.989  & 0.988  \\
LSTM    & 0.741    & 0.184     & 0.500  & 0.269  \\ \hline
\end{tabular}
\caption{Scores over the test dataset. N = normalized, L = Lower, P = no position embedding.}
\label{tab:scores}
\end{table}

@@

The following inputs has been tagged with the CNN P model. Batch are constructed around the regex \texttt{\\W} with package \texttt{regex}. This explains why inputs such as \texttt{".i."} are automatically tagged as \texttt{" . i . "} by the tool. The input was stripped of its spaces before tagging, we only show the ground truth by commodity.

@@

\section{Annex 1 : Confusion of CNN without position embeddings}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{confusion.png}
  \caption{Confusion matrix of the CNN model without position embedding}
  \label{fig:confusion_matrix}
\end{figure}
